---
title: "Poisson regression for rate data"
author: "Michael D. Garber, PhD MPH"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r "setup", include=FALSE, eval=TRUE, echo=FALSE}
require("knitr")
library(here)
```

# Outline

1.  Overview of Poisson regression for rate data
2.  Load British Doctor's Data
3.  Fit the model in the British Doctor's Data
4.  Use the model to predict values

# Poisson regression for modeling rates

A rate is defined as the number of events divided by some denominator that measures the at-risk experience giving rise to those events. In epidemiology, it is often expressed as number of events per person time.

Note this is different from a "risk", because a risk can be viewed as a probability bounded by 0 and 1, whereas a rate is not bounded by 1.

We can use poisson regression to model those rates as a function of predictor variables.

Specifically, we model a Poisson distributed variable (integer 0, 1, 2, 3, ...) (<https://en.wikipedia.org/wiki/Poisson_distribution>), and we include the denominator of the rate as an "offset."

Technically, we model the logarithm of the rate as a linear function of predictor variables.

As described in video 9.9 below, the model takes the following functional form:

ln(rate)=ln(y/t) = b0+b1\*x1+b2\*x2 +...

where

-   y is the number of events (or a count),

-   t is follow-up time (the rate's denominator),

-   b0 is the intercept,

-   b1 is the coefficient for variable x1,

-   b2 is the cofficient forindependent variable x2,

and additional variables and coefficients can be added.

Then, because of the properties of logarithms, we can rewrite the log(rate) as:

ln(y/t)=ln(y)-ln(t)

So the full model can be expressed as

ln(rate)=ln(y/t) =ln(y)-ln(t)= b0+b1\*x1+b2\*x2

Now we can move ln(t) over to the other side

ln(y)-ln(t)= b0+b1\*x1+b2\*x2

ln(y)= b0+b1\*x1+b2\*x2 +ln(t)

This ln(t) is what is known as the "offset" when fitting the Poisson model for rate data in software.

## Suggested YouTube videos on Poisson regression

I recommend this series of YouTube videos by MarinStatsLectures-R Programming, which discuss theory and application of the Poisson model

-   9.9 Poisson Regression: The Model For Rate Data (what is an offset?)
    <https://www.youtube.com/watch?v=5pZbER_mR4k&list=RDCMUCaNIxVagLhqupvUiDK01Mgg&index=4>


-   9.10 Poisson Regression in R: Fitting a Model To Rate Data (with offset) in R
    <https://www.youtube.com/watch?v=QP4F98ysrEA>


-   9.11 Poisson Regression: Model Assumptions
    <https://www.youtube.com/watch?v=r492fF9XZl8&t=2s>


The videos use the British doctor's data, a famous study on lung cancer and smoking: <https://en.wikipedia.org/wiki/British_Doctors_Study>Load British Doctor's Data

To download British doctor's data, we can use this package: <https://rdrr.io/cran/ACSWR/man/bs1.html>

```{r,eval=FALSE}
install.packages("ACSWR") #This installs the package containing the data
```

```{r}
library(ACSWR) #This loads the package containing the data
data(bs1) #This brings data into the working environment.
bs1 #Take a look at the data.
```

I like to work in the tidyverse, so let's make it a tibble (a data frame with some nice formatting).

Please see here for an intro to data wrangling in the tidyverse: <https://michaeldgarber.github.io/teach-r/dplyr-1-nyt-covid.html>

```{r}
library(tidyverse)

bs1_tibble=bs1 %>% as_tibble()

bs1_tibble
```

In looking at the data, we can see that the variable `Deaths` in this dataset is the Poisson (count) variable.

And the denominator for the rate is `Person_Years`.

So the rate is rate=(Deaths/Person_Years)

# Fit a Poisson regression to model the rate

Following video 9.10 above, let's model the rate as a function of smoking (yes/no) and the age category.

We use the `glm()` function to fit this model.

```{r}
#Use names() to get variable names to make sure we type them correctly in the model.
names(bs1_tibble)

pois_model = glm(
  formula=  Deaths~ #  #the outcome variable is Deaths, a count variable
    Smoke_Ind+Age_Group+ #the two predictor variables
    
    #here is where we specify the offset, i.e., the denominator of the rate
    offset(log(Person_Years)),
  
  #Poisson family for Poisson regression. Note lowercase. R is case-sensitive.
  family="poisson",   
  data=bs1_tibble #the data in which the model is fit
           )

#Note this can all be done on fewer lines, as long as arguments are separated by commas.
#I put on multiple lines so I could annotate each.
pois_model = glm(
  formula=  Deaths~ Smoke_Ind+Age_Group+ offset(log(Person_Years)),
  family="poisson", data=bs1_tibble
  )

```

Check out the model results by printing the object corresponding to the model.

```{r}
pois_model
```

The rate ratio corresponding to each predictor can be found by exponentiating the coefficient. The coefficients can be accessed by the syntax `model_object$coefficients`

```{r}
pois_model_coefficients=pois_model$coefficients
rate_ratios=exp(pois_model_coefficients)
rate_ratios
```

So adjusting for age category, smokers have a 1.4-times higher rate of lung cancer death.

# Use the model to predict values

We can use this model to predict expected values based on values of age category, smoking, and the number of person-years.

We use the `predict.glm()` to do this. Each argument is described in this documentation: <https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/predict.glm>

Below, we specify the following arguments: `object`, `newdata`, and `type` .

The `object` is the object corresponding to the model that we fit.

`newdata` is the dataset in which the values will be predicted. This can be any dataset as long as it has the variables included in the model.

If we say, `type="response"`, the resulting predicted value will be on the scale of the response variable.

```{r}
predicted_values=predict.glm(
  pois_model,#the model,
  
  #We can use any dataset in "newdata" as long as it has the variables included in the model.
  #Here, we're simply going to use the original data in which the model was fit.
  #We can then compare the true values in the dataset with the model's predicted values for that observation.
  newdata = bs1_tibble,
  type="response"
) %>% 
  as_tibble()#Convert the result to a tibble for easier data wrangling
```

We can see here that we have just one column of data. These are the expected values corresponding to each row in the original dataset.

```{r}
predicted_values
```

We can append these predicted values to the original dataset using the dplyr function `bind_cols().` I'm creating a new tibble called `bs1_tibble_w_pred_values` which begins with `bs1_tibble` and then gets `predicted_values` appended to it.

```{r}
bs1_tibble_w_pred_values=bs1_tibble %>% 
  bind_cols(predicted_values) %>% 
  #The predicted values are called "value". Let's change that to 
  #Deaths_expected
  rename(Deaths_expected=value) %>% 
  #We could calculate a residual
  #Residual=actual value - predicted value
  mutate(residual=Deaths-Deaths_expected)

```

Check out the data including the predicted values:

```{r}
bs1_tibble_w_pred_values
```

Now let's do some checks.

## Print actual and expected number of deaths for each observation

We can examine the actual number of deaths and the expected number based on the model to see how well the model performed.
Here I'm using dplyr::select() to just select the two deaths variables and then will print them in the console so we can compare the two.

```{r}

bs1_tibble_w_pred_values %>% 
  dplyr::select(Deaths,Deaths_expected)
```

## Histogram for residuals

How are the residuals distributed?

```{r}
bs1_tibble_w_pred_values %>% 
  ggplot()+
  geom_histogram(
    aes(residual),#the variable to be visualized with the histogram
    bins = 5#customize the histogram's bin width
    )
```


